---
title:       "Methods of Process Modelling"
author:      "Max Arthur Hachemeister"
date:        last-modified
date-format: iso
format:
  gfm:
    toc:     true
---

# Prerequisites

```{r}
#| label: setup
#| results: false

library(tidyverse)
library(tidymodels)
theme_set(theme_light())
```

# The task

## Technical

-   6 Pages Minimum

## Content

-   Compare the the observed growth of *Picea abies* of two different locations
-   Select and fit appropriate forest growth models
    -   Visualize the growth at both locations and compare
-   Present the fitted models
    -   With respect to their coefficients
-   Select a suitable allometric model to calculate
    -   Above-ground Biomass (average)
    -   120 year old tree
    -   For each location
-   Reflect and describe *precisely* the reasons for selecting the allometric models
    -   Range, sample size, coefficient of determination
    -   Present the respective results
-   Interpret and discuss the results with respect to the differences between the two locations
    -   stand composition, climate conditions, elevation, etc.
-   Create a document, following scientific writing principles

## Remark

> It is not necessary to document every processing step of the data and/or your script when working through the task. You may name and cite the respective software and/or packages.
>
> Concentrate on explaining and introducing the selected approach as well as most important settings, on the results, their presentation (tables, figures, etc.), and interpretation.
>
> Do not forget to give references and transparently note when you used further tools.

# The data

## Data dictionary

+------------+--------------------------------------------------------------------------------+
| Variable   | Description                                                                    |
+============+================================================================================+
| location   | 1 = cold temperate climate, altitude \> 1000 m above sea level                 |
|            |                                                                                |
|            | 2 = temperate climate, altitude \< 1000 m above sea level (e.g. Harz, Germany) |
+------------+--------------------------------------------------------------------------------+
| tree       | ID of observed *Picea abies* tree                                              |
+------------+--------------------------------------------------------------------------------+
| age.base   | **Age** of respective tree \[years\]                                           |
+------------+--------------------------------------------------------------------------------+
| height     | Measured **stem height** of respective tree and age \[m\]                      |
+------------+--------------------------------------------------------------------------------+
| dbh.cm     | Measured stem **diameter at breast height** \[cm\]                             |
+------------+--------------------------------------------------------------------------------+
| volume     | calculated **stem volume** per tree and age \[m³\]                             |
+------------+--------------------------------------------------------------------------------+

# Fumbling around

## Import

```{r}
trees <- 
  read_delim("data/examdata.txt", 
             delim = "\t",
             escape_double = FALSE, 
             trim_ws = TRUE,
             col_types = "ffiddd"
             )

trees <- 
  janitor::clean_names(trees) |> 
  rename(age = age_base)

trees
```

# Explore
## Summary
Always nice!

```{r}

trees |> 
  summary()
```

There are more observations for location 7, however, the two trees at location 2 have the most respective observations, which in this case just means that they have been observed for the longest total time period. Let's check that:

```{r}
# Check the max age
trees |> 
  group_by(tree) |>
  summarise(n = n(),
            min_age = min(age),
            max_age = max(age)
            )
```

## Common tree height

Good, all trees have been measured first with 20 years, and the minimal `max_age` is 80 let's compare the trees at that age:
```{r}
trees |> 
  filter(age == 80)
```
Location seems to explain the most of the differences in the other variables, interesting.

## Volume calculations

Let's check how the volume got to be, because this was calculated, but the equation was not given in the data dictionary.

Let's start with the basic forestry equation of $\frac{\pi}{4} d^2 \times h \times f$, with $f = 1/2$

```{r}
# make it a function
eq_volume_forestry <- function(diameter, height, factor = 1/2){
  pi/4 * diameter * height * factor
}

# calculate the volume
trees |> 
  mutate(
    volume_test = round(eq_volume_forestry(dbh_cm, height), 2)
    ) |> 
  select(tree, volume, volume_test)
```
The original volume values seem quite high.
Let's compare that visually:
```{r}
trees |> 
  mutate(
    volume_test = round(eq_volume_forestry(dbh_cm, height), 2)
    ) |> 
  pivot_longer(contains("volume")) |> 
  ggplot(aes(age, value, color = location)) +
  geom_point() + 
  geom_smooth(aes(group = location),
              method = "loess")+
  facet_wrap(~ name) +
  labs(
    title = "Comparison of Different Volume Calculations",
    subtitle = "Volume_test was calculated with the standard forestry equation.",
    caption = "The results differ strongly.",
    x = "Age",
    y = NULL
  ) +
  scale_y_continuous(labels = scales::label_number(suffix = "m³")) +
  scale_x_continuous(labels = scales::label_number(suffix = "y"))
```
Yep, that's quite a difference. I became sceptical of calculations I did not do in my own scripts.
Let's consider for now, that the original `volume` values are unreliable.

## Visualize

```{r}
trees |> 
  pivot_longer(height:dbh_cm,
               names_to = "variable",
               values_to = "value"
               ) |> 
  ggplot(aes(age,
             value,
             group = tree, 
             color = location
             )
         ) +
  geom_point() +
  geom_smooth(aes(group = location),
              method = "loess",
              se = TRUE
              )+
  facet_wrap(~ variable,
             scales = "free"
             )
```
