---
title:       "Methods of Process Modelling"
author:      "Max Arthur Hachemeister"
date:        last-modified
date-format: iso
execute:
  echo: false
format:
  gfm:
    toc:     true
bibliography: references.bib
---

# Prerequisites

```{r}
#| label: setup
#| results: false

library(tidyverse)
library(tidymodels)
library(growthrates)
theme_set(theme_light())
```

# The task

## Technical

-   6 Pages Minimum

## Content

-   Compare the the observed growth of *Picea abies* of two different locations
-   Select and fit appropriate forest growth models
    -   Visualize the growth at both locations and compare
    -   Present the fitted models
        -   With respect to their coefficients
-   Select a suitable allometric model to calculate
    -   Above-ground Biomass (average)
    -   120 year old tree
    -   For each location
-   Reflect and describe *precisely* the reasons for selecting the allometric models
    -   Range, sample size, coefficient of determination
    -   Present the respective results
-   Interpret and discuss the results with respect to the differences between the two locations
    -   stand composition, climate conditions, elevation, etc.
-   Create a document, following scientific writing principles

## Remark

> It is not necessary to document every processing step of the data and/or your script when working through the task. You may name and cite the respective software and/or packages.
>
> Concentrate on explaining and introducing the selected approach as well as most important settings, on the results, their presentation (tables, figures, etc.), and interpretation.
>
> Do not forget to give references and transparently note when you used further tools.

# The data

## Data dictionary

+------------+--------------------------------------------------------------------------------+
| Variable   | Description                                                                    |
+============+================================================================================+
| location   | 2 = cold temperate climate, altitude \> 1000 m above sea level                 |
|            |                                                                                |
|            | 7 = temperate climate, altitude \< 1000 m above sea level (e.g. Harz, Germany) |
+------------+--------------------------------------------------------------------------------+
| tree       | ID of observed *Picea abies* tree                                              |
+------------+--------------------------------------------------------------------------------+
| age.base   | **Age** of respective tree \[years\]                                           |
+------------+--------------------------------------------------------------------------------+
| height     | Measured **stem height** of respective tree and age \[m\]                      |
+------------+--------------------------------------------------------------------------------+
| dbh.cm     | Measured stem **diameter at breast height** \[cm\]                             |
+------------+--------------------------------------------------------------------------------+
| volume     | calculated **stem volume** per tree and age \[m³\]                             |
+------------+--------------------------------------------------------------------------------+

# Fumbling around

## Import

```{r}
trees <- 
  read_delim("data/examdata.txt", 
             delim = "\t",
             escape_double = FALSE, 
             trim_ws = TRUE,
             col_types = "ffiddd"
             )

trees <- 
  janitor::clean_names(trees) |> 
  rename(age = age_base)

trees
```

# Explore

## Summary

Always nice!

```{r}

trees |> 
  summary()
```

There are more observations for location 7, however, the two trees at location 2 have the most respective observations, which in this case just means that they have been observed for the longest total time period. Let's check that:

```{r}
# Check the max age
trees |> 
  group_by(tree) |>
  summarise(n = n(),
            min_age = min(age),
            max_age = max(age)
            )
```

## Common tree height

Good, all trees have been measured first with 20 years, and the minimal `max_age` is 80 let's compare the trees at that age:

```{r}
trees |> 
  filter(age == 80)
```

Location seems to explain the most of the differences in the other variables, interesting.

## Volume calculations

Let's check how the volume got to be, because this was calculated, but the equation was not given in the data dictionary.

Let's start with the basic forestry equation of $m^3 =\frac{\pi}{4} d^2 \cdot h \cdot f \cdot 0.001$, with $f = 0.5$

```{r}
# make it a function
volume_forestry <- function(diameter, height, factor = 0.5, digits = 3){
  result <- 
    (pi/4 * diameter^2) * height * factor * 1e-4
  
  
  round(result, digits = digits)
}

# calculate the volume
trees |> 
  mutate(
    volume_test = volume_forestry(dbh_cm, height, .5)
    )
```

The original volume values seem quite high for cubic meters. Maybe it's cubic or centimeters, or decimeters?

I guess it's decimeters, so adjusting the original volume values by 0.001:

```{r}
trees |> 
  mutate(
    volume_test = round(volume_forestry(dbh_cm, height), 2),
    volume_cdm = volume * 1e-3
    ) |> 
  pivot_longer(contains("volume_")) |> 
  ggplot(aes(age, value, color = location)) +
  geom_point() + 
  facet_wrap(~ name) +
  labs(
    title = "Comparison of Different Volume Calculations",
    subtitle = "Volume_test was calculated with the standard forestry equation.",
    # caption = "The results differ strongly.",
    x = "Age",
    y = NULL
  ) +
  scale_y_continuous(labels = scales::label_number(suffix = "m³")) +
  scale_x_continuous(labels = scales::label_number(suffix = "y"))
```

Yep, something like that, with a correction factor of 0.001. Still some difference! I became skeptical of calculations I did not do in my own scripts. Let's consider for now, that the original `volume` values might be right, but the actual equation is unknown so the values are unreliable.

Basically it is not known which *volume* in particular was calculated. Ah, now that I look at the data dictionary it's the *stem volume*.

The equation above is the standard for merchantable wood; i.e., every part of a tree where $d >= 7cm$. Let's try the $DENZIN$ [@kramer2008]equation and then move on.

The equation is:

$$
V = 2(d^{2}/1000) \cdot (h - NH) \cdot Volumenkorrekturfaktor
$$

where:

-   $d = \text{DBH, [cm]}$
-   $h = \text{Total tree height, [m]}$
-   $NH = \text{"Normalhöhe", aka. height of a standard tree}$
-   $\text{Volumenkorrekturfaktor} = \text{Reduction factor}$

```{r}
volume_denzin_spruce <- function(dbh_cm, height_m, formfactor = 0.04, digits = 2){
  result <- 
    (dbh_cm^2 / 1000) + (dbh_cm^2 / 1000) * (height_m - (19 + 2 * (dbh_cm * .1))) * formfactor 
  
  round(result, digits = digits)
}

trees |> 
  mutate(
    volume_denzin = volume_denzin_spruce(dbh_cm, height),
    volume_forest = volume_forestry(dbh_cm, height),
    volume_dcm = round(volume * 1e-3, 2)
  ) |> 
  select(1:2, 7:9)
```

Gotcha, maybe some rounding divergence. Ahja I didn't round down to full centimeters. Let's do that and see:

```{r}
trees |> 
  mutate(
    volume_denzin = volume_denzin_spruce(floor(dbh_cm), height),
    volume_forest = volume_forestry(dbh_cm, height),
    volume_dcm = round(volume * 1e-3, 2)
  ) |> 
  select(1:2, 7:9)

```

Still a little off, so let's stick with $DENZIN$ and rounding down, so that I at least know how the values got to be:

```{r}
trees <- 
  trees |> 
    mutate(volume = volume_denzin_spruce(dbh_cm = floor(dbh_cm),
                                       height_m = height,
                                       digits = 2
                                       )
           )
```

## Visualize

Let's compare the two locations visually. I let simple linear regressions be fitted for each location:

```{r}
trees |> 
  pivot_longer(height:volume,
               names_to = "variable",
               values_to = "value"
               ) |> 
  ggplot(aes(age,
             value,
             group = tree, 
             color = location
             )
         ) +
  geom_point() +
  geom_smooth(aes(group = location),
              method = "lm",
              formula = y ~ x,
              se = TRUE
              )+
  facet_wrap(~ variable,
             scales = "free"
             )
```

Okay, looks like the location is a rather significant predictor for the size in our data. Let's see how location and tree would be employed by a linear regression:

```{r}
lm(volume ~ location + tree, data = trees) |> 
  summary() |> 
  coef()
```

Yes, the both locations - `location 2` being the `intercept` - are strong predictors for the tree volume, according to the data we have; which is not much.

# Forest Growth Models

Since the locations affect the trees growth, the data set is split to fit the models respectively.

```{r}
trees_l2 <- 
  trees |> 
  filter(location == 2)

trees_l7 <- 
  trees |> 
  filter(location != 2)
  
```

### Define Function and Parameters

```{r}

x_l2       <- trees_l2$age
x_l7       <- trees_l7$age
y0         <- .01
mumax      <- .2
beta       <- .2 # for richards

parameters_l2 <- c(
  y0 = y0,
  mumax = mumax,
  K = max(trees$height),
  beta = beta
)

parameters_l7 <- c(
  y0 = y0,
  mumax = mumax,
  K = max(trees$height),
  beta = beta
)


y1_l2         <- grow_logistic(x_l2, parameters_l2) |> as_tibble(y)
y2_l2         <- grow_gompertz(x_l2, parameters_l2) |> as_tibble(y)
y3_l2         <- grow_richards(x_l2, parameters_l2) |> as_tibble(y)

y1_l7         <- grow_logistic(x_l7, parameters_l7) |> as_tibble(y)
y2_l7         <- grow_gompertz(x_l7, parameters_l7) |> as_tibble(y)
y3_l7         <- grow_richards(x_l7, parameters_l7) |> as_tibble(y)
```

### Model calibration

```{r}
#| warning: false

# Fit location 2
fit1_l2 <- fit_growthmodel(grow_logistic,
                        p = parameters_l2,
                        time = trees_l2$age,
                        y = trees_l2$height)

fit2_l2 <- fit_growthmodel(grow_gompertz,
                        p = parameters_l2,
                        time = trees_l2$age,
                        y = trees_l2$height)

fit3_l2 <- fit_growthmodel(grow_richards,
                        p = parameters_l2,
                        time = trees_l2$age,
                        y = trees_l2$height)

# Fit location 7
fit1_l7 <- fit_growthmodel(grow_logistic,
                        p = parameters_l7,
                        time = trees_l7$age,
                        y = trees_l7$height)


fit2_l7 <- fit_growthmodel(grow_gompertz,
                        p = parameters_l7,
                        time = trees_l7$age,
                        y = trees_l7$height)

fit3_l7 <- fit_growthmodel(grow_richards,
                        p = parameters_l7,
                        time = trees_l7$age,
                        y = trees_l7$height)

trees_rename <- 
trees |> 
  rename(
    time = age,
    value = height
  )

trees_logistic <- 
  all_growthmodels(
    value ~ grow_logistic(time, parms) | location,
    data = trees_rename,
    p = parameters_l2
  )

trees_gompertz <-
  all_growthmodels(
    value ~ grow_gompertz(time, parms) | location,
    data = trees_rename,
    p = parameters_l2
  )

trees_richards <- 
    all_growthmodels(
    value ~ grow_richards(time, parms) | location,
    data = trees_rename,
    p = parameters_l2
  )



map(list(trees_gompertz, trees_logistic), \(x) deviance(x))

list_c(predict(trees_gompertz)) |> as_tibble() |> 
  select(height_gompertz = 2)
                                   
```

### Check the results

### Residuals

#### Prepare rownames

```{r}
# Check residuals
rownames <- c("logistic", "gompertz", "richards")

```

#### Squared Sum of Residuals

```{r}
residuals_l2 <- c(sum(fit1_l2@fit$residuals^2),
                sum(fit2_l2@fit$residuals^2),
                sum(fit3_l2@fit$residuals^2))

residuals_l7 <- c(sum(fit1_l7@fit$residuals^2),
                sum(fit2_l7@fit$residuals^2),
                sum(fit3_l7@fit$residuals^2))

residuals <- 
  tibble(model = rownames,
         l2 = residuals_l2,
         l7= residuals_l7,
         )

residuals |> 
arrange(l2, l7)

deviance(fit1_l2)

```

#### R Squared

```{r}
# Check rsquared
rsquared_l2 <- c(fit1_l2@rsquared,
                fit2_l2@rsquared, 
                fit3_l2@rsquared)

rsquared_l7 <- c(fit1_l7@rsquared,
                fit2_l7@rsquared, 
                fit3_l7@rsquared)

rsquared <- 
  bind_cols(list(
       l2 = rsquared_l2,
       l7 =rsquared_l7
       )
  ) |> 
  mutate(model = rownames,
         .before = 1)


rsquared |> 
  arrange(across(c(!model),desc))
    
```

#### Predict new values and plot

```{r}
# predict values for each model
res1_l2 <- as_tibble(predict(fit1_l2))
res2_l2 <- as_tibble(predict(fit2_l2))
res3_l2 <- as_tibble(predict(fit3_l2))

res1_l7 <- as_tibble(predict(fit1_l7))
res2_l7 <- as_tibble(predict(fit2_l7))
res3_l7 <- as_tibble(predict(fit3_l7))

model_results_list <- 
  list(res1_l2, res2_l2, res3_l2, res1_l7, res2_l7, res3_l7)

#combine
model_results <- 
bind_rows(model_results_list, .id = "id") |> 
  mutate(
    location = as_factor(
      if_else(id <= 3, 2, 7)
      ),
    model = as_factor(
      case_when(
        id %in% c(1, 4)  ~ "logistic",
        id %in% c(2, 5) ~ "gompertz",
        TRUE ~ "richards"
        )
      )
    )

# plot
model_results |> 
  ggplot() +
  geom_point(data = trees,
             aes(age, height),
             color = "grey") +
  geom_line(aes(time, y, color = model),
            linewidth = 2/3) +
  facet_wrap(~location) +
  scale_color_discrete(palette = "set2")
```

## Allometric models

-   Select a suitable allometric model to calculate
    -   Above-ground Biomass (average)
    -   120 year old tree
    -   For each location

### Define the model and its parameters

We're interested in the (average) Above Ground Biomass, and the date above has shown, that the locations are quite different, therefore the models should be diligently selected.

What do we know:

-   Species: Picea abies
-   Location 7: Probably Germany
-   Location 2: \>1000m above sea level, and cold temperate
    -   according to the wikipedia for both: austria
        -   but no AB model for that
        -   Elevation, norway, sweden maybe, but climate bit too cold
        -   Czec republic, fits both reasonably well.
-   Diameter and Height ranges

```{r}
trees |> 
  group_by(location) |> 
  summarize(
    min_diameter = min(dbh_cm),
    max_diameter = max(dbh_cm),
    min_height = min(height),
    max_height = max(height)
  )
```

Both Czech are probably the same sample just either with meter, or not, so take both:

-   141
-   142

Germany, let's see:

-   147
    -   does not have the minimum range, but the max range is fine; also has the most samples, and the only R² mention

Let's define the models and their standard parameter values:

```{r}
model_cz_141 <-
  function(diameter, a = 0.57669 , b = 1.964){
  
    a * diameter^b
  }

model_cz_142 <- 
  function(diameter, height, a = 0.11975, b = 0.81336){
  
    a * ((diameter^2) * height)^b
  }

model_de_147 <-
  function(diameter, a = -43.13, b = 2.25, c = 0.452){
  
    a + b * diameter + c * diameter^2
  }
```

And apply those models to the dataset:

It should be noted that $AB = \text{Above Ground Biomass [kg]}$

```{r}
allometric_models <- 
  list(
    model_cz_141,
    model_cz_142,
    model_de_147
  )

trees_allometric <-  
  trees |> 
    mutate(
      AB_141 = if_else(location == 2, model_cz_141(dbh_cm), NA),
      AB_142 = if_else(location == 2, model_cz_142(dbh_cm, height), NA),
      AB_147 = if_else(location == 7, model_de_147(dbh_cm, height), NA)
    )

trees_allometric |> 
  group_by(location, age) |> 
  summarize(
    across(contains("AB"), mean),
    .groups = "drop"
  ) |> 
  filter(age == 120)
```

### Plot it also, maybe:

```{r}

allometric_pivot <- 
  trees_allometric |> 
    pivot_longer(cols = contains("AB"),
                 names_to = "model",
                 values_to = "AB") |> 
  filter(!is.na(AB))
  
allometric_pivot |> 
  ggplot(aes(age, AB, color = model)) +
  geom_point(color = "grey") +
  geom_point(data = allometric_pivot |>  filter(age == 120),
             shape = 1,
             size = 2,
             stroke = 1) +
  ggrepel::geom_text_repel(data = allometric_pivot |> filter(age == 120),
            aes(age, AB, label = paste(round(AB, 2), "kg")),
            nudge_x = - 60,
            direction = "y"
            # nudge_y = -300
            ) +
  facet_wrap(~location) +
  scale_color_discrete(palette = "set2")
```

### TODO Maybe model the one missing tree with his 120 years.

Not all trees have been measured at 120 years. We can use the growth models we fit to predict those values though:

Ah, no we can't yet because the models were fit for height only. So I would nee to fit another model for diameter.

```{r}
fit3_l2_diameter <- fit_growthmodel(grow_richards,
                        p = parameters_l2,
                        time = trees_l2$age,
                        y = trees_l2$dbh_cm)

tidy_model_info(x = fit3_l2_diameter)
  


tidy_model_info <- function(x) {
  
  model_name <- deparse(substitute(x))
  
  tibble(
    model = model_name,
    residuals_sqsum = rsquared(x),
    r_squared = sum(residuals(x)^2)
  )
  
}
```
